---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# cattonum

[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/cattonum)](https://cran.r-project.org/package=cattonum)
[![Build Status](https://travis-ci.org/bfgray3/cattonum.svg?branch=develop)](https://travis-ci.org/bfgray3/cattonum)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/bfgray3/cattonum?branch=develop&svg=true)](https://ci.appveyor.com/project/bfgray3/cattonum)
[![Coverage status](https://codecov.io/gh/bfgray3/cattonum/branch/develop/graph/badge.svg)](https://codecov.io/github/bfgray3/cattonum?branch=develop)
[![Total Downloads](http://cranlogs.r-pkg.org/badges/grand-total/cattonum)](https://cran.r-project.org/package=cattonum)
[![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
[![Licence](https://img.shields.io/github/license/mashape/apistatus.svg)](http://choosealicense.com/licenses/mit/)

## Special note

I am currently looking for someone who would be interested in becoming the official maintainer of the package. Advanced skill in R or previous package development experience is not necessary. I will be able to help out with code review, etc.  Please comment on [this issue](https://github.com/bfgray3/cattonum/issues/40) if you are interested.

## Summary

cattonum ("cat to num") provides different ways to encode categorical features as numerics.  Its goal is to be a one-stop shop for all categorical encoding needs.  It includes the following:

* dummy encoding: `catto_dummy()`
* frequency encoding: `catto_freq()`
* label encoding: `catto_label()`
* leave-one-out encoding: `catto_loo()`
* mean encoding: `catto_mean()`
* median encoding: `catto_median()`
* one-hot encoding: `catto_onehot()`

There are many existing packages with which to encode categorical features, including (among others):

* [CatEncoders](https://cran.r-project.org/package=CatEncoders)
* [dummies](https://cran.r-project.org/package=dummies)
* [embed](https://github.com/tidymodels/embed)
* [fastDummies](https://cran.r-project.org/package=fastDummies)
* [FeatureHashing](https://cran.r-project.org/package=FeatureHashing)
* [h2o](https://cran.r-project.org/package=h2o)
* [makedummies](https://cran.r-project.org/package=makedummies)
* [recipes](https://cran.r-project.org/package=recipes)
* [vtreat](https://cran.r-project.org/package=vtreat)

## Installation

The development version can be installed from GitHub.

```{r, eval = FALSE}
remotes::install_github("bfgray3/cattonum", ref = "develop")
```

The latest official release can be installed from CRAN.

```{r, eval = FALSE}
install.packages("cattonum")
```

## Usage

```{r}
library(cattonum)
data(iris)
head(catto_loo(iris, response = Sepal.Length))
```

## Code of Conduct

Please note that the 'cattonum' project is released with a [Contributor Code of Conduct](.github/CODE_OF_CONDUCT.md).  By contributing to this project, you agree to abide by its terms.
